\textit{MSF: Modulated Sub-graph Finder}

### 2018-05-07 MSF_Paper_FA.tex
editing of several parts

High throughput techniques such as RNA-seq or microarray analysis have
proven to be invaluable for the characterization of global transcriptional
gene activity changes due to external stimuli or diseases. Differential
gene expression analysis (DGEA) is the first step in the course of data
interpretation, typically producing lists of of dozens to thousands of
differentially expressed genes. To further guide the interpretation of
these lists, different pathway analysis approaches have been
developed. These tools typically classify genes into set of genes, i.e. a
pathway, which interact with each other and function in a common biological
process. Regardless of technical differences, all these methods do not
properly account for cross talk between different pathways and rely on
binary classification into differentially expressed gene and unaffected
genes based on an arbitrarily set p-value cut-off.

@@

Existing pathway-based analysis approaches use different research
designs, which can be categorized into ORA (Over-representation
analysis), FCS (Functional class scoring) and pathway topology based
methods. All of which aim in finding in a subset of genes, e.g.,
significantly differentially expressed genes, genes associated with a
certain pathway more often than expected given the total set of
examined genes, e.g.~the whole genome background.  ORA is the first
and the most basic method of pathway analysis. It uses a DEG list with
user defined cut-off for the log-fold change and \textit{p}-value
(most commonly using absolute log-fold change $\geq$ 2 and
\textit{p}-value $\leq$ 0.05). Subsequently, sets of genes associated
with annotated pathways are tested for being over-represented in the
set of DEGs. To this end, hyper-geometric distribution, chi-square
tests, binomial probability or the Fisher’s exact test are
used. Thereby the information of the topology of genes in the pathways
are ignored~\cite{Bayer}. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they
cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

As input, information on functional links between genes provided by
e.g.~KEGG or Reactome and information on the differential expression
status of single genes resulting from a DGEA, are required. As a
result the analysis returns sub-graphs and their joint confidence
scores, reflecting how the perturbation is migrated through the
network. Furthermore, the entry points of perturbation in the networks
and overlap with conventional pathway categories are returned. The
output is prepared in a directed adjacency file, convenient for
visualization with the StringApp~\cite{StringApp}, available as a
Cytoscape plug-in~\cite{Cyto}.

@@

\texttt{MSF} uses the individual gene's \textit{p}-values generated
from the DGEA. The \textit{p}-value expresses the probability that the
null hypothesis of unmodified gene expression can be rejected for a
given statistical model. To find significantly modulated sub-graphs
individual \textit{p}-values of the vicinal genes in the global
network are combined into a single combined \textit{p}-value, using a
statistical method for combining dependent \textit{p}-values described
by Hartung~\cite{Hartung}. Hartung uses the inverse of standard normal
distribution function, individual gene \textit{p}-values are first
transformed to their corresponding normal score. Then using these
normal scores, the correlation between genes is calculated, the normal
scores and correlation are applied to the inverse normal function to
calculate the combined \textit{p}-value for all genes examined, namely
the examined sub-graph. The combined \textit{p}-value of a sub-graph
will express the significance of all genes in the sub-graph being
modulated together. Thereby, the information from the different genes
are used as, although not independent, replicated measurement of the
behavior of the whole sub-graph. This potentially increases the power
to detect also significant sub-graphs consisting of genes which are
not significant on there own.
\newline

@@

In the next step initial modulated sub-graphs are used to check if
they could further be extended beyond the immediate neighborhood. This
is done by testing all possible extension paths up to N genes for all
genes in the sub-graph. Again, this step is iteratively repeated until
no further genes are added to the significant differentially expressed
sub-graphs. This steps bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

@@

After detection and extension of the modulated sub-graphs, they are tested
if combined sub-graphs score better than on their own. The merging of the
two sub-graphs is done by depth first search. If the two sub-graphs merge
with the connector of at most N genes (default two genes) and the combined
\textit{p}-value of the merged sub-graph including the bridging genes in
between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

@@

To demonstrate its usefulness, \texttt{MSF} is applied to an RNAseq
data set of primary human monocyte-derived macrophages (MDMs) infected
with Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs
to the Filoviridea family; filamentous, enveloped and single stranded
RNA viruses. EBOV causes hemorrhagic fever in humans, inducing the
host innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has
been proposed that IFN-$\alpha / \beta$ should be tested against Ebola
for its antiviral activity through clinical trials~\cite{Konde}. Ebola
infection data was selected to test the approach because it has been
well recognized in the literature for the last several decades. It
would be easy to cross verify the results of \texttt{MSF} with the
vast literature present on Ebola infection.  The aim of testing this
data with \texttt{MSF} was to identify the modulated sub-graphs and
check if \texttt{MSF} is able to identify the IFN-$\alpha / \beta$
gene as one of the sources, and show that the capability to close
``gaps'' by using information of neighboring genes is effective.

@@

The EBOV infection
experiments describe the course of infection at three time-points 6,
24 and 48 hour post infection (hpi). For the earliest time point at
6~hpi, five modulated sub-graphs were identified with 41, 107, and 18
genes. Two of the five sub-graphs were less than 4 genes long. Most of
the genes part of the sub-graphs were cytokines, chemokines (CXCL10,
CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as a subset
of CD4+ T helper cells. IFNB1 and IFNA1 were both identified as two of
the possible sources. Most of the sources identified by \texttt{MSF}
were type~I interferon induced genes. At 24~hpi nine modulated
sub-graphs were identified with four main sub-graphs consisting of 38,
85, 148 and 167 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point
48~hpi, eleven modulated sub-graphs were identified. Eight of the
sub-graphs were less than seven genes and main sub-graphs had 39, 71
and 210 genes. IFNB1 and IFNA1 were identified as the two sources out
of the possible sources.

As stated earlier IFN-$\alpha / \beta$ could be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1
as a source in all three time-points and INFB1 in two of the
time-points. Although IFNA1 and IFNB1 were two of the most significant
gene in the later time points, \texttt{MSF} was able to detect them as
a source in the early time-point when the genes were not significant
based on the individual DGEA alone. Identifying the possible sources
will reduce the search space for potential target genes and can help
the biologist as the starting point of clinical testing for drugs and
vaccines against an infection.

Table~\ref{tab:rawVsHartung} shows if \texttt{MSF} does not
consider the complete DEG results and uses different cut offs for
significant genes from DEG results, how the signal transduction would
fall apart into incomprehensible small sub-graphs and single unconnected
genes.

@@

 This show cast example might advertise with how little effort complex
 data cat be interpreted, help to apprehend the dynamics of the
 underlying processes and suggest testable hypothesis and potential
 points of intervention.

@@

 To assess the robustness and stability of our method, Poisson
 distributed noise was added to the read counts of the three
 time-points data set, used in the previous analysis. Then DGEA was
 carried out on the disturbed data with the same parameters as for the
 native data using edgeR, followed by analysis with \texttt{MSF}. This
 procedure was carried out 100 times.  Every time the genes from the
 modulated sub-graphs identified from noisy data were compared to the
 genes of sub-graphs identified from the native data. This was also
 done for the DEG obtained for each run, a cutoff of FDR less than
 0.05 was used to compare the retrieved DEG from noisy data to the
 native data.  The robustness of \texttt{MSF} and the DEG analysis for
 the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can
 be considered as rather stringent, which is already reflected limited
 recall rate in the edgeR based DGEA, between 84~\% (6~hpi) and 93~\% (24~hpi). For
 \texttt{MSF}-analysis the observed median recall rates lay between
 71~\% (6~hpi) and 84~\% (48~hpi).

@@

  Gene enrichment analysis was performed on \texttt{MSF} identified sub-graphs for each time-point using Reactome Analyze data tool~\cite{Reactome}. Reactome's over-representation analysis tests whether certain Reactome pathways are enriched by the list of genes submitted. For comparison the DEG results from edgeR were filtered using a cut off of adjusted p-value < 0.1. This subset of DEG list was used for gene enrichment analysis. Enriched pathways were further filtered by confidence score p-values, pathways having p-value less $<$ 0.05 for cut off DEG list and $>$ 1$^{-10}$ for  pathway enrichment in \texttt{MSF} identified sub-graph genes. Out of 282 pathways dis-regulated in the two comparison analysis with three time-points each, pathways that were present in at least three time-points are shown in Fig~\ref{fig:msfvsreactome}.

  The comparison shows most of the important pathways known from literature to be dis-regulated by Ebola infection are enriched in both the enrichment analysis. Toll-Like receptor signaling pathway when interacts with EBOV glycoprotein (GP), triggers the activation of cytokines~\cite{Olejnik}. This pathway was not identified when a p-value cut off DEG list was analyzed for enrichment. Due to \texttt{MSF} considering the complete DEG results, even the weak signal at the earliest time-point was detected; for example Toll-like receptor signaling.

@@

Classic pathway analysis tools aim to detect in lists of significantly
deregulated genes enriched associations with pathway genes categorized
by their biological function and their interactions. Thereby,
depending on the tool, the internal pathway topology is considered or
neglected all together. The here presented tool, \texttt{MSF}, employs
a different approach, by aiming to detect sub-graphs in whole gene
regulatory networks which are significantly deregulated in a concerted
manner. To this end, neighboring genes in the user provided network
are tested for jointly common regulation. Exploiting that each gene's
abundance, although not independent from its neighbors, is measured
repeatedly on its own, sensitivity can be increased by our applied
p-value meta-analysis, namely Hartung's method. This potentially
enables to call just not significant modulated genes based on the DGEA
to be convincingly called to be part of a deregulated gene group.
Furthermore, it allows to identify connected sub-graphs, representing
the propagation of gene regulation perturbation in the input
network. A better understanding of this propagation, especially the
critical spots such as sensors, effectors, or intermediate bottlenecks
and hubs, facilitates the projection of potential intervention points,
e.g. for drug development. Since \texttt{MSF} only uses interaction
information in gene regulation network, but not the functional
grouping of the genes into functional pathways, it is especially
adapted to discover so called cross-talk between such pathways.

@@

\texttt{MSF} is a fast, robust and easy to use tool to find
concertedly modulated sub-graphs in a given network. Its
implementation in java enables it use across many operating
systems. It requires as input the results from a differential gene
expression analysis in the appropriate file format. So far the raw
output from edgeR~\cite{edgeR} and DESeq2~\cite{love2014moderated} are supported
. Furthermore, a gene network has to be provided in the format of a directed adjacency list.

@@

The Ebola infection RNAseq dataset analyzed during the current study
are available in the GEO repository (GSE84188)~\cite{Olejnik}. The cell signaling network file used is from Reactome
Functional interactions (FIs) Version 2016~\cite{Cytokegg}

@@

Source code: https://github.com/MariamFarman/Modulated-SubGraph-Finder \newline
Software license: Creative Commons Attribution 4.0 International License
\section*{Author contributions}
Conceptualization: FA, ILH; Funding Acquisition: ILH; Resources: ILH;
Software: MRF; Writing – Original Draft Preparation: MRF; Writing –
Review \& Editing: FA, ILH. All authors read and approved the final
manuscript.

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{17 cm}
			\includegraphics[width=16cm]{DEGvsMSF2.png}
			\caption{The percentage of DEG analysis genes
                          and MSF identified sub-graph genes recall
                          rates for the three different time points of
                          EBOV infection data for 100 simulation where
                          Poisson distributed noise was added to the
                          experimental deduced reads per gene
                          counts.\label{fig:DEGvsMSF}}
			\label{label}
		\end{minipage}
	}
\end{figure*}

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{10 cm}
	\includegraphics[width=1.0\linewidth]{MSFvsReactome}
	\caption{The heatmap shows Reactome pathway enrichment for MSF identifed sub-graphs for the three time-points and the Reactome pathway enrichment for subset of DEG with adjusted p-value cut off of < 0.1. Green color shows the pathway being enriched and red color shows pathway not found as enriched.}
	\label{fig:msfvsreactome}
\end{minipage}
}
\end{figure*}

>>>>

High throughput techniques such as RNA-seq or microarray analysis have
proven to be invaluable for the characterization of global transcriptional
gene activity changes due to external stimuli or diseases. Differential
gene expression analysis (DGEA) is the first step in the course of data
interpretation, typically producing lists of dozens to thousands of
differentially expressed genes. To further guide the interpretation of
these lists, different pathway analysis approaches have been
developed. These tools typically rely on the classification of genes into
set of genes, i.e. a pathway, based on the interactions between the genes
and their function in a common biological process. Regardless of technical
differences, these methods do not properly account for cross talk between
different pathways and rely on binary separation into differentially
expressed gene and unaffected genes based on an arbitrarily set p-value
cut-off.

@@

Existing pathway-based analysis approaches use different research designs,
which can be categorized into ORA (Over-representation analysis), FCS
(Functional class scoring) and pathway topology based methods. All of which
aim in finding in a subset of genes, e.g., significantly differentially
expressed genes, genes associated with a certain pathway more often than
expected given the total set of examined genes, e.g.~the whole genome
background.  ORA is the first and the most basic method of pathway
analysis. It uses a DEG list with user defined cut-off for the log-fold
change and \textit{p}-value (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
associated with annotated pathways are tested for being over-represented in
the set of DEGs. To this end, hyper-geometric distribution, chi-square
tests, binomial probability or the Fisher’s exact test are used. Thereby
the information of the topology of genes in the pathways is
neglected~\cite{Bayer}. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they
cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

As input, information on functional links between genes provided by
e.g.~KEGG or Reactome and information on the differential expression status
of single genes resulting from a DGEA, are required. As a result the
analysis returns sub-graphs and their joint confidence scores, reflecting
how the perturbation is migrated through the network. Furthermore, the
entry points of perturbation in the networks and overlap with conventional
pathway categories are returned. The output is prepared in a directed
adjacency file, convenient for visualization, e.g., with
StringApp~\cite{StringApp}, available as a Cytoscape plug-in~\cite{Cyto}.

@@

\texttt{MSF} uses the individual gene's \textit{p}-values generated from
the DGEA. The \textit{p}-value expresses the probability that the null
hypothesis of unmodified gene expression can't be rejected for a given
statistical model. To find significantly modulated sub-graphs individual
\textit{p}-values of the vicinal genes in the global network are combined
into a single combined \textit{p}-value, using a statistical method for
combining dependent \textit{p}-values described by
Hartung~\cite{Hartung}. Hartung's method uses the inverse of standard
normal distribution function, individual gene \textit{p}-values are first
transformed to their corresponding normal score. Then using these normal
scores, the correlation between genes is calculated, the normal scores and
correlation are applied to the inverse normal function to calculate the
combined \textit{p}-value for all genes examined, namely the examined
sub-graph. The combined \textit{p}-value of a sub-graph will express the
significance of all genes in the sub-graph being modulated
together. Thereby, the information from the different genes are used as,
although not independent, replicated measurement of the behavior of the
whole sub-graph. This potentially increases the power to detect also
significant sub-graphs consisting of genes which are not significant on
there own.
\newline

@@

In the next step initial modulated sub-graphs are used to check if they
could further be extended beyond the immediate neighborhood. This is done
by testing all possible extension paths up to \emph{N} genes for all genes
in the sub-graph. Again, this step is iteratively repeated until no further
genes are added to the significant differentially expressed
sub-graphs. This step bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

@@

After detection and extension of the modulated sub-graphs, they are tested
if combined sub-graphs score better than on their own. The merging of the
two sub-graphs is done by depth first search. If the two sub-graphs merge
with the connector of at most \emph{N} genes (default two genes) and the
combined \textit{p}-value of the merged sub-graph including the bridging
genes in between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

@@

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized
in the literature for the last several decades \TODO{what has been
  recognized??}.  Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
rightness and usefulness of the approach.

@@

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, five modulated sub-graphs were identified with 41, 107, and
18 genes. Two of the five sub-graphs were less than 4 genes long. Most of
the genes part of the sub-graphs were cytokines, chemokines (CXCL10, CCL8)
and Interleukin genes (IL6, IL27, IL23) which serve as a subset \TODO{??}
of CD4+ T helper cells. IFNB1 and IFNA1 were both identified as two of the
possible sources. Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes. At 24~hpi nine modulated sub-graphs were
identified with four main \TODO{(more than X genes)} sub-graphs consisting
of 38, 85, 148 and 167 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point 48~hpi,
eleven modulated sub-graphs were identified. Eight of the sub-graphs were
less than seven genes and main sub-graphs had 39, 71 and 210 genes. IFNB1
and IFNA1 were identified as the two sources out of the possible sources
\TODO{??}.

As stated earlier IFN-$\alpha / \beta$ was reported to be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1 as a
source in all three time-points and INFB1 in two of the
time-points. Although IFNA1 and IFNB1 were already among the most
significant gene in the DGEA during the later time points, \texttt{MSF} was
also able to detect them as a source in the very early time-point when the
genes were not significant based on the individual DGEA alone. Identifying
the possible sources will reduce the search space for potential target
genes and can help the biologist as the starting point of clinical testing
for drugs and vaccines against an infection.

Table~\ref{tab:rawVsHartung} compares the results of \texttt{MSF}, namely
the number of detected sub-modules and their total genes numbers, to a
simple analysis of mapping significantly modulated genes from the DGEA to
the network and joining neighbors to modules. The numbers indicate that
\texttt{MSF} detects less but larger sub-modules, applying its statistical
test.  Furthermore, the dependency of the results from the p-value cutoff
choice is demonstrated for the DGEA, which is avoided for \texttt{MSF}
altogether.

@@

 This show cast example might advertise with how little effort complex data
 can be interpreted, help to apprehend the dynamics of the underlying
 processes and suggest testable hypothesis and potential points of
 intervention.

@@

 To assess the robustness and stability of our method, Poisson distributed
 noise was added to the read counts of the three time-points data set, used
 in the previous analysis. Then DGEA was carried out on the disturbed data
 with the same parameters as for the native data using edgeR, followed by
 analysis with \texttt{MSF}. This procedure was carried out 100 times.
 Every time the genes from the modulated sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the native
 data. This was also done for the DEG obtained for each run, a cutoff of
 FDR less than 0.05 was used to compare the retrieved DEG from noisy data
 to the native data.  The robustness of \texttt{MSF} and the DEG analysis
 for the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
 considered as rather stringent, which is already reflected by the limited
 recall rate in the edgeR based DGEA, between 84~\% (6~hpi) and 93~\%
 (24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
 between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
 pure DGEA can be explained by the fact that disturbed p-values do not
 change the results for DGEA as long as the p-value does not wise above the
 chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
 changes across the whole range of possible values.

@@

Gene enrichment analysis was performed on genes in \texttt{MSF} identified
sub-graphs for each time-point using Reactome Analyze data
tool~\cite{Reactome}. Reactome's over-representation analysis tests whether
certain Reactome pathways are enriched by the list of genes submitted. For
comparison the DEG results from edgeR were filtered using a cut off of
adjusted p-value < 0.1. This subset of DEG list was used for gene
enrichment analysis. Enriched pathways were further filtered by confidence
score p-values, pathways having p-value less $<$ 0.05 for cut off DEG list
and $>$ 1$^{-10}$ for pathway enrichment in \texttt{MSF} identified
sub-graph genes. Out of 282 pathways dis-regulated in the two comparison
analysis with three time-points each, pathways that were present in at
least three time-points are shown in Fig~\ref{fig:msfvsreactome} \TODO{rephrase; not really clear how this filtering of pathways shown was done}.

The comparison shows most of the pathways known from literature
to be dis-regulated by Ebola infection are enriched in both the enrichment
analysis. Toll-Like receptor signaling pathway when interacts with EBOV
glycoprotein (GP), triggers the activation of
cytokines~\cite{Olejnik}. This pathway was not identified when a p-value
cut off DEG list was analyzed for enrichment. Due to \texttt{MSF}
considering the complete DEG results, even the weak signal at the earliest
time-point was detected; for example Toll-like receptor signaling. \TODO{what is the point you want to say and state it clearly. this is to vague}

@@

Classic pathway analysis tools aim to detect in lists of significantly
deregulated genes enriched associations with pathway genes categorized by
their biological function and their interactions. Thereby, depending on the
tool, the internal pathway topology is considered or neglected all
together. The here presented tool, \texttt{MSF}, employs a different
approach, by aiming to detect sub-graphs in whole gene regulatory networks
which are significantly deregulated in a concerted manner. To this end,
neighboring genes in the user provided network are tested for jointly
common regulation. Exploiting that each gene's abundance, although not
independent from its neighbors, is measured repeatedly on its own,
sensitivity can be increased by our applied p-value meta-analysis, namely
Hartung's method. This potentially enables to call just not significant
modulated genes based on the DGEA to be convincingly called to be part of a
deregulated gene group.  Furthermore, it allows to identify connected
sub-graphs, representing the propagation of gene regulation perturbation in
the input network. A better understanding of this propagation, especially
the critical spots such as sensors, effectors, or intermediate bottlenecks
and hubs, facilitates the projection of potential intervention points,
e.g., for drug development. Since \texttt{MSF} only uses interaction
information in gene regulation network, but not the functional grouping of
the genes into functional pathways, it is especially adapted to discover so
called cross-talk between such pathways.

@@

\texttt{MSF} is a fast, robust and easy to use tool to find concertedly
modulated sub-graphs in a given network. Its implementation in java enables
its use across many operating systems. It requires as input the results from
a differential gene expression analysis in the appropriate file format. So
far the raw output from edgeR~\cite{edgeR} and
DESeq2~\cite{love2014moderated} are supported. Furthermore, a gene network
has to be provided in the format of a directed adjacency list.

@@

The Ebola infection RNA-seq dataset analyzed during the current study are
available in the GEO repository (GSE84188)~\cite{Olejnik}. The cell
signaling network file used is from Reactome Functional interactions (FIs)
Version 2016~\cite{Cytokegg}

@@

Source code:
https://github.com/MariamFarman/Modulated-SubGraph-Finder \newline Software
license: Creative Commons Attribution 4.0 International License

\section*{Author contributions}
Conceptualization: FA, ILH; Funding Acquisition: ILH; Resources: ILH;
Software: MRF; Writing – Original Draft Preparation: MRF; Writing – Review
\& Editing: FA, ILH. All authors read and approved the final manuscript.

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{17 cm}
			\includegraphics[width=16cm]{DEGvsMSF2.png}
			\caption{\TODO{make plot smaller and consider to add DEG vor 0.1, 0.05 and 0.01 as you did before} The percentage of DEG analysis genes
                          and MSF identified sub-graph genes recall
                          rates for the three different time points of
                          EBOV infection data for 100 simulation where
                          Poisson distributed noise was added to the
                          experimental deduced reads per gene
                          counts.\label{fig:DEGvsMSF}}
			\label{label}
		\end{minipage}
	}
\end{figure*}

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{10 cm}
	\includegraphics[width=1.0\linewidth]{MSFvsReactome}
	\caption{The heatmap shows Reactome pathway enrichment for genes in \texttt{MSF} identifed sub-graphs for the three time-points and the Reactome pathway enrichment for subset of DEG with adjusted p-value cut off of < 0.1. Green color shows the pathway being enriched and red color shows pathway not found as enriched. \TODO{this is not really a heatmap (no quantitative value encoded in the color but only qualitative value); you might want to include different p-value cutof for the DGE; in your set up you have to comapre column 1 with column 4; column 2 with column 5, and so on; that is suboptimal; the time points are not chronological ordered.}}
	\label{fig:msfvsreactome}
\end{minipage}
}
\end{figure*}
